---
title: "Introduction to R and RStudio: Key"
subtitle: "BIOST 311, Discussion Section Week 1"
author: "Taylor Okonek and Charlie Wolock"
date: "March 29th, 2022"
output: 
  pdf_document:
    toc: true
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)

## function to color text
colFmt = function(x,color){
  outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
```

## R: the big picture

R is a `r colFmt("free", "blue")`, `r colFmt("open source", "orange")` software package that can be used for data analysis, graphics, and programming.

At its core, R is an interactive, command-driven `r colFmt("language", "blue")`: you type a command, R executes the command, and then returns results.

While it may take some time to develop a mastery of R, it is relatively easy to get set up with the basics and start analyzing data.

Learning R is worth your time, and has many advantages, including:

  * R is free and open source
  * there is an active group of contributors to R software
  * you can do flexible things with your data
  * many people have written packages that make data analysis easier
  
To begin using R, you need to download and install it. You can download the latest version of R [here](https://www.r-project.org/).

## RStudio: the interface to R

While R is the `r colFmt("language", "blue")` you will use to analyze data, it does not come packaged with the best `r colFmt("interface", "orange")`. An interface connects you to the program: for example, your operating system lets you point and click on icons (and read things easily) rather than writing in 0's and 1's, the language that the computer understands.

Instead of using the R interface packaged with the download, you will use RStudio. This is also a `r colFmt("free", "blue")`, `r colFmt("open source", "orange")` software package, but is only an interface to R---this means that even if you type into RStudio, the commands will be executed in R. Technically, RStudio is called an integrated development environment (IDE).

RStudio adds many things over the basic R interface, including:

  * An improved layout of tools for data analysis
  * Support for embedding reproducible research tools into data analysis
  * Support for building your own R packages
  * Integrated R help and documentation
  * De-bugging tools
  
The combination of R and RStudio makes it easy to:

  * write R scripts containing all code for an analysis, along with comments describing what the code does
  * write reports with code embedded (in Rmarkdown, more on this later)
  * organize your data analysis workflow (e.g., reading in data, accessing help files)
  
As with R, you need to download RStudio before you can get started. You can download RStudio [here](https://www.r-project.org/).

When you open RStudio for the first time, the window will display three `r colFmt("panels", "blue")`---these have different functions, as you'll now discover.

### The console

The default position of the console is in the lower left-hand pane of RStudio. This pane is your real window into R: any commands that you run in R will appear in the console, and any results that are output by these commands, with the exception of plots, will appear in the console. 

To enter a command into the console, move your cursor to the `>` and click. When there is a flashing `|` in the console, you're ready to enter a command. From now on, you'll call the `>` the execution line. Once you've entered a command on the execution line, hit `Enter` to run the command.

For example, you could type in the number `47` and hit enter, which prints out the number `47`:
```{r example-47, include = TRUE, echo = TRUE}
47
```

Or, you could use R as a calculator:
```{r example-calculator, include = TRUE, echo = TRUE}
32 + 15
52 - 5
94/2
23*2 # This is a comment! Everything after a '#' isn't run by R; instead, it's printed out. Here, you could say that the result isn't 47!
```

Each of these lines returns an `r colFmt("object", "blue")`---in each case, the object is the result of the computation. In fact, these objects are called `r colFmt("vectors", "blue")`. Vectors are the building block of R objects.

### The script editor

It is possible to do all of your work in the console. However, this isn't advisable, because:

* you may forget the commands you've run, and
* you may forget the objects you've created;

both of these (and other reasons) make `r colFmt("reproducibility", "blue")` difficult if you only use the console. 

Reproducibility is important both for yourself and for others: 

* you should always structure your code so that future you can come back in 6 months and figure out what you are doing
* you may have to run your code more than once, so why not write it down only once?
* someone else should be able to take your code and data and obtain the same results as you did

The `r colFmt("script editor", "blue")` is one place to write down and save your code. By default, it is located in the upper left panel of RStudio. This pane serves as a text editor; it allows you to edit and save code and Rmarkdown documents, among other things.

There are two primary ways to write and save R code: RMarkdown files and R files. 

To create a new RMarkdown file, click on `File > New File > R Markdown`. RMarkdown files are saved with the file extension `.Rmd`. These files allow you to seamlessly integrate code with text (this document is written in Rmarkdown!), and also include graphs, plots, tables, etc. You can then output your RMarkdown file to Word (.docx), PDF (.pdf), or HTML (.html). You create the output by clicking `Knit`.

To create an R script file, click on `File > New File > R Script`. These files, saved with the file extension `.R`, allow you to save R code and execute it in the console. They are simpler than `.Rmd` files, but also much more limited. One advantage of using RMarkdown over R scripts is that you can write text explaining what your code does!

To run code from the source editor, use `ctrl + Enter` (PC) or `cmd + Enter` (Mac), which allows you to either run a single line (whichever line your cursor is on); or multiple lines, by highlighting multiple lines. (You can also use your mouse by clicking `Run > Run Selected Lines`, but this quickly becomes tedious.)

Practice running the following line of code from your .Rmd file into the console:
```{r example-run-line, echo = TRUE, include = TRUE}
45 + 2
```

The places in the .Rmd file that begin and end with three back-ticks are called `r colFmt("code chunks", "blue")`. This is where you enter any code that you want to be executed by R when you knit your .Rmd file. You can also test your code by running the code from the chunks in the console. Use `ctrl + shift + Enter` or `cmd + shift + enter` to run an entire chunk of code at once. 

### The environment/history pane

The environment/history pane is, by default, located in the upper right-hand panel of RStudio. The `r colFmt("Environment", "orange")` tab shows any functions or data that you have in your `r colFmt("workspace", "blue")`; that is, any objects created by running R code. Up to now, your workspace only includes the function `colFmt()`, which allows me to color text in Rmarkdown.

The `r colFmt("History", "orange")` tab shows all of the commands that have been entered into the console. 

While the environment/history panel can sometimes be helpful, it is `r colFmt("not", "red")` a substitute for using R scripts or Rmarkdown!

### The files/plots/packages/help pane

The final panel, in the bottom right, by default hosts files/plots/packages/help. This is where any plots that you generate will be displayed, and any help files that you access will appear (more on both of these later!).

## R: commands

R commands are the lines of code you run; they relate `r colFmt("functions", "blue")` to `r colFmt("objects", "orange")`. Functions `r colFmt("do", "blue")` things, while objects `r colFmt("store", "orange")` values. Everything in R is done by a function---for example, the code you entered in before ran the `print()` function. Both of the following commands return the same value:
```{r example-print, include = TRUE, echo = TRUE}
45 + 2
print(45 + 2)
```

### Functions

Functions take in `r colFmt("arguments", "blue")`. This is how you, the user, tell a function what to do. Some functions just perform an action, and others return a `r colFmt("value", "orange")`. A function is accessed by typing its name, followed by an open and closed set of parentheses: for example, `read.csv()` is a function that can read in data in a comma-separated file (.csv). 

`r colFmt("Arguments", "blue")` to functions go between the parentheses, and are separated by commas. You specify what values the arguments should take on by using `=`; for example, `read.csv(file = "births.csv", header = TRUE)` reads in the data in `births.csv`, and treats the first line of the file as a header (containing column names, not actual data).

Functions return `r colFmt("values", "orange")`, which are generally a combination of objects, plots, and printouts. Printouts just appear on your console, while objects and plots can be given names and stored for later. The value of `read.csv(file = "births.csv", header = TRUE)` is a special R object called a `data.frame`, which is essentially equivalent to a dataset.

### Creating objects

R objects are (basically) the result of calling functions. However, there are two general ways this can be done:

* `r colFmt("loading data", "blue")` (e.g., from a .txt or .csv file)
* `r colFmt("manipulating another object", "orange")` using a function

You assign a value to an object using the special character `<-`. The value, on the right-hand side, can now be accessed using its object name (on the left-hand side).

I suggest naming your objects so that the name gives you some information about the object---don't go too overboard, but this practice has saved me lots of time. For example, `births` is a much more informative name for a dataset than `x`. 

### Reading in a dataset

Our example for today is the King County births data, which you'll see many times this quarter. First, you need to read the data into R. In order to do this, you need to tell R where the data file lives on your computer.

A `r colFmt("working directory", "blue")` is where R will read files from, and also save files (like plots, if you choose). You might create a folder on your computer for your homework assignments or your final project, and store all relevant files there. So set your working directory during an R session, use hte `setwd()` command. I'll set my working directory right now to where the data is on my computer --- yours will not be the same!

After we've told R where it can find the data, we can read the data file.

```{r read-data, echo = TRUE, include = TRUE}
## load the FEV data
setwd("C:/Users/cwolo/Dropbox/UW/TA/BIOST311/Datasets/")
births <- read.csv(file = "births.csv", 
                header = TRUE)
## check that you've read it in correctly:
head(births)
```

What happened here? First, the `setwd()` command set my working directory to the folder that holds all my datasets for this course. Then, the `read.csv()` function read in the data; you then used `<-` to assign those data, as a value, to the object called `births`. Finally, you used the `head()` function to print the first six rows of the `births` object.

## Data structures

Reading a dataset into R isn't the end of the story---often, you'll need to compute `r colFmt("descriptive statistics", "blue")`, `r colFmt("plot", "orange")` certain variables, and `r colFmt("run statistical analyses", "cyan")`, sometimes on only part of the data. Throughout this quarter, you'll learn tools to do all of these things. 

I mentioned earlier that vectors are the building block of R objects---vectors are 1-dimensional, so they only have a length. You can create a vector using the `c()` function:

```{r create-vector, echo = TRUE, include = TRUE}
my_vector <- c(1,2,3,4,5)
print(my_vector)
```

`r colFmt("Data frames", "blue")`, R's way of storing datasets, are collections of vectors: each column of a data frame is a `r colFmt("variable", "orange")`, which is stored as a vector; similarly, each row of a data frame is an `r colFmt("observation", "red")`, which is also stored as a vector. 

You access values in vectors using open and closed square brackets `[]`. For example, let's say I want to access the third element in `my_vector`. I can do this by telling R to pull out the third element via `my_vector[3]`. Be careful - if you give an index that doesn't exist in the vector, you'll get `NA`. (This is an example of when R lets you make a mistake without telling you. Learning to watch out for these instances will become more intuitive with practice.) 

```{r access-vector, echo = TRUE, include = TRUE}
my_vector[3] # this is fine
my_vector[6] # this returns NA
```

You can access multiple values using the `c()` function.

```{r access-multiple-vector, echo = TRUE, include = TRUE}
my_vector[1,3]
```

You access values in data frames using the same square brackets---however, since data frames have rows and columns, you access a value in the `i`th row and `j`th column using `[i, j]`. 

For example, access the 3rd element in the 5th column of the births data:
```{r example-access-value, echo = TRUE, include = TRUE}
births[c(3, 5)]
```

From your earlier call to the `head()` function, you'll notice that this output corresponds to the 3rd observation on the variable `parity`.

## R packages

Since R is open source software, many people contribute to its development. This comes in the form of `r colFmt("packages", "blue")`: bundles of functions that supplement the functions always loaded into R. 

You will use several packages extensively throughout this course. We'll start with the `tidyverse` package, which is very useful for manipulating data. 

The first time you want to use a package, you'll have to install it. Most packages can be installed using `install.packages("package name")`. 

Once the package is installed, each subsequent time you want to use functions from the package, you have to load it using `library`---and you have to do this `r colFmt("each time", "red")` you close and re-open R or RStudio.

For example, load the `tidyverse` package:
```{r example-package-load, echo = TRUE, include = TRUE}
# run this next line (without the #) if you haven't installed the package yet
# install.packages("tidyverse")
# run this next line every time you open a new R session
library("tidyverse")
```

## Data wrangling

Using the `tidyverse` package, you can pass the output of one function directly to another --- that way, you don't have to save all the intermediate objects as separate variables. This keeps things much cleaner! To pass output from one function to the next, using the `%>%` operator (referred to as "pipe.") 

For example, suppose we want to take the `age` column of the births dataset, and then look at some rows. We can pick columns using `select()`. The `head()` function will show the first 6 rows, while we can choose arbitrary chunk of rows using `slice()`. (Our input to `slice()` is `20:30`, which simply lists all integers from 20 to 30.) The `%>%` operator links these commands together. 

```{r example-select, echo = TRUE, include = TRUE}
## get the names of the data frame
names(births)
## look at just the age column
births %>% select(age) %>% head()
## access the 20th through 30th values of age
births %>% select(age) %>% slice(20:30)
```

The `names()` function returns the variable names in a data frame. 

The final major concept in manipulating datasets is `r colFmt("logical expressions", "blue")`. The two logical values are `TRUE` and `FALSE`; we can make comparisons that will return these values. For example, is 3 > 5?
```{r example-logical-comp, echo = TRUE, include = TRUE}
3 > 5
```

You can now `r colFmt("subset", "orange")` your data by making logical comparisons. For example, you might be interested only in those study participants who are less than or equal to 20 years old. This means you're interested in `r colFmt("all columns", "blue")` from the births data, but only for the participants (i.e., `r colFmt("observations", "red")`) whose `age` value is `<= 20`. We perform subsetting based on logical expressions using the `filter` function. 

```{r example-subset, echo = TRUE, include = TRUE}
## subset the data, keep all columns
under20 <- births %>% filter(age <= 20)
## simple summary statistics
summary(under20)
```

The `summary()` function you just used displays simple descriptive statistics. This is a great way to get a quick snapshot of the data, and make sure they have been read in correctly. Look at the descriptive statistics for `wgain`. Do you notice anything interest about this variable? What are its largest and smallest values?

*One interesting thing about the `wgain` variable is that the minimum value in the dataset is exactly 0. This should make us ask whether or not it's possible to have a negative value (i.e., how would the data be recorded if someone lost weight during pregnancy?). In fact, if we were to look more closely, we'd see that 19 individuals in this dataset had recorded weight gain of exactly 0. To me, this suggests that negative weight gain was recorded as 0.*

### Creating a new variable

You can also use the `tidyverse` package to create new variables using the `mutate()` function. A baby is considered preterm if it is born at 36 weeks gestation or earlier. We'll create a new binary variable called `preterm` that is 1 for preterm births and 0 otherwise. We need to use the `<-` symbol to save our altered dataset to the variable called `births`. 

```{r mutate, echo = TRUE, include = TRUE}
births <- births %>% mutate(preterm = ifelse(gestation <= 36, 1, 0))
```

We used `ifelse()` to create the new binary variable: this takes a logical comparison (e.g., is `gestation` less than or equal to 36?), what to assign if the expression is true (e.g., `1`), and what to assign if the expression is false (e.g., `0`).

*This does not, however, save these variables into the underlying .csv file: for that, you'll need a different function, `write.csv()`*. 

## Getting help

Until now, you've had functions provided to you, with the relevant arguments filled out. But what if you don't know what arguments a function has? 

Typing `?` in the console, followed by the function name, calls up the help file for the function. All R functions are required to have help files.

```{r example-help, echo = TRUE, include = TRUE}
?summary
```

If you don't know the name of the function, you aren't out of luck. If you know what you'd like to do, you can use `??`, followed by a one-word description, to pull up all help files that mention the word.

```{r example-help-2, echo = TRUE, include = TRUE}
??summarize
```

Finally, Google, Stack Exchange, Canvas, and your friends are your best resource: if you have an R problem, I can almost guarantee that someone else has had it before, and the answer is either on the internet or with your friend!

## Conclusion

You're now on your journey to becoming true data scientists! 

Today, you've learned that R is a powerful software package that gives you flexible options for analyzing your data. You've also learned how to do simple arithmetic, read in a dataset, create objects, add variables to data sets, calculate simple descriptive statistics, access help files, and load packages. 

While it may be frustrating at times, learning R is worth your time. Additionally, using RMarkdown and RStudio is a great first step to making sure that all of your analyses are reproducible---and being fully reproducible is important for science!