---
title: "Homework 1: R cheat sheet"
subtitle: "BIOST 311"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
```

This cheat sheet will walk you through some `R` commands that will be useful as you complete Homework 1. 

We will rely heavily on the `tidyverse` package, so let's start by loading it.

```{r load_tidy}
library(tidyverse)
```

We'll also load the `births` dataset to use as an example. (Note that the function we use here reads the data directly from the course GitHub page; you can also load it from your computer like we did in Week 1 Discussion Section. If you want to do that, change this line of code!)

```{r load_data, results = "hide", message = FALSE, error = FALSE, warning = FALSE}
births <- readr::read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/births.csv")) # change this if you want to load from your computer
```

## Subsetting

There are two basic functions for subsetting data: `filter` and `select`. The former picks out \textcolor{blue}{rows} of the data frame, while the latter picks out \textcolor{blue}{columns}. 

Usually, we use `filter` by giving it a logical expression, in order to pick out rows meeting some criteria. For example, to get all observations from the dataset where the birth parent smoked cigarettes, we can run

```{r filter_smoke}
births_smoke <- births %>% filter(smoke == 1) # recall what the pipe operator (%>%) does!
```

The basic logical operators are

* `==` equal to 
* `!=` not equal to
* `>` greater than
* `<` less than
* `>=` greater than or equal to
* `<=` less than or equal to

Note that we only have only 175 individuals in this dataset now, after filtering out all those who did not smoke. (We can count rows of the dataset using `nrow`.)

```{r nrow}
nrow(births_smoke)
```

We can filter according to multiple criteria using the `&` ("and") and `|` ("or") symbols. 

```{r and_or}
births_and <- births %>% filter(smoke == 1 & drink == 1) # those who smoked and drank
births_or <- births %>% filter(smoke == 1 | drink == 1) # those who smoked or drank
```

We will often want to pull out particular columns of a data frame, usually to pass on to other functions. We can pick one or multiple columns using `select`.

```{r select_1}
births %>% select(race) %>% head() # look at first six rows
```

```{r select_2}
births %>% select(race, gestation) %>% head() # look at first six rows
```

## New variables

We may want to create new variables from the existing variables in our dataset. We use the `mutate` function for this. If you give the new variable the same name as an existing variable in your dataset, it will overwrite the existing variable. Be careful with this!

We can create a new variable that's the logarithm of `bwt`. Note that we use the `<-` symbol to assign the output of our function to an object. If we don't do that, the output will just appear in our console without being stored anywhere.

```{r mutate}
births %>% mutate(l_bwt = log(bwt)) # not very useful
births <- births %>% mutate(l_bwt = log(bwt)) # now our new variable is stored in our data frame
```

A common operation used in creating a new variable is `ifelse`. This functions takes three arguments: (1) a logical expression; (2) a value to assign if that logical expression is `TRUE`; (3) a value to assign if that logical expression is `FALSE`. For example, say we want to create a new binary variable indicator whether an individual had more than 12 years of education. We can use `ifelse` to do this quite simply: 

```{r ifelse}
births <- births %>% mutate(edu12 = ifelse(education > 12, 1, 0))
```

If you want to convince yourself that this worked, you can manually look at a few observations to see if individuals with `education > 12` got assigned a value of 1, as we expect. 

```{r confirm}
births %>% select(education, edu12) %>% head()
```

## Descriptive statistics

We will spend a lot of time in BIOST 311 looking at descriptive statistics -- it's the first step in any data analysis. `R` has a decent built-in function for basic summary statistics, aptly named `summary`. Let's look at `age`, `race`, and `parity`.

```{r summary_function}
births %>% select(age, race, parity) %>% summary()
```

Since `race` is a string (i.e., composed of alphanumeric characters), `summary` doesn't tell us much. We can learn a lot more by creating a `factor` version of `race`. A `factor` is an `R` data type that represents a categorical variable (either ordinal or nominal). Let's create a `factor` version of `race` and try `summary` again. 

```{r summary_factor}
births %>% mutate(race_f = factor(race)) %>% select(race_f) %>% summary()
```

Now that we have a factor, `summary` actually tells us something useful: how many individuals fall into each category. 

We can also perform custom descriptive statistics using the `tidyverse` function `summarize`. Let's say we want the standard deviation of `bwt`, which isn't given to us by `summary`. We can use the `sd` (standard deviation) function inside of `summarize`:

```{r summarize}
births %>% summarize(bwt_sd = sd(bwt))
```

We can also get multiple summary statistics at the same time. Let's look at the interquartile range (IQR) along with the standard deviation. 

```{r summarize_multiple}
births %>% summarize(bwt_sd = sd(bwt), bwt_iqr = IQR(bwt))
```
