---
title: "Homework 1: Key"
subtitle: "BIOST 311"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = "hide")
```


## Instructions

This assignment should be submitted on Canvas as a .pdf by 12:30pm on Friday, April 8th. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code (if applicable) in the same document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

Each homework assignment is worth 20 points total. Problems are graded for correctness, according to the number of points given in the problem statement. 

Be sure to look at the cheat sheet for this week's homework -- it will provide you with useful `R` functions!

## 1. Descriptive statistics and data visualization

We will be using the `births` dataset often in the first few weeks of BIOST311. For this question, you will get acquainted with some of the variables in the dataset. 


```{r preliminaries}
library(readr)
library(tidyverse)
library(ggplot2)
library(knitr)
```

a. Load the `births` dataset (0 points)

```{r 1a}
births <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/births.csv"))
```

b. The `gestation` variable gives the number of weeks gestation for each birth in the dataset. Give relevant descriptive statistics for this variable, and provide one plot of your choice to visualize its distribution. Would you describe the distribution as roughly symmetric, left skewed, or right skewed? (2 points)

\textcolor{blue}{Descriptive statistics for the gestation times (in weeks) are given in the table below. We chose a standard five number summary (using the `fivenum` function in `R`), which includes minimum, first quartile, median, third quartile, and maximum. Another reasonable option for a continuous variable like `gestation` would be a mean and standard deviation.}

```{r 1b1, results = "markup"}
desc <- births %>% pull(gestation) %>% fivenum()
table_df <- data.frame("Min." = desc[1],
                       "Q1" = desc[2],
                       "Med." = desc[3],
                       "Q3" = desc[4],
                       "Max." = desc[5])
kable(table_df)
```

\textcolor{blue}{We also include a boxplot of the gestation times. A histogram would be another reasonable choice.}

```{r 1b2, results = "markup"}
births %>% ggplot(aes(x = gestation)) + 
  geom_boxplot() + 
  theme_bw() + 
  xlab("gestation (weeks)") + 
  ggtitle("Boxplot of gestation times") + 
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())
```

c. Make another plot of gestation times, but this time stratified by birth parent's smoking status. Qualitatively describe how the two distributions look in comparison to one another. (2 points)

\textcolor{blue}{The figure below is a boxplot of gestation times stratified by smoking status. For looking at quantitative variables stratified by group, side-by-side boxplots are generally the cleanest option. Interlaced or overlapping histograms could work, but are generally a bit more difficult to analyze visually.}

```{r 1c, results = "markup"}
births <- births %>% mutate(smokef = factor(smoke, levels = c(0, 1), labels = c("No", "Yes")))
births %>% ggplot(aes(x = gestation, y = smokef)) + 
  geom_boxplot() + 
  theme_bw() + 
  xlab("gestation (weeks)") + 
  ylab("smoking status") + 
  ggtitle("Boxplot of gestation times by smoking status")
```

\textcolor{blue}{Overall, the distributions of gestation times for people who smoke vs. those who do not look quite similar. The distribution for those who smoke is a bit more left-skewed, with a lower first quartile. The centers of each distribution are nearly identical, as are the ranges.}

d. The `wpre` variables gives the weight (in pounds) of the birth parent prior to pregnancy, and `wgain` gives the weight gained during pregnancy. Make a single plot examining the relationship between these two variables. Qualitatively describe the association between `wpre` and `wgain`. (2 points)

```{r 1d}
births %>% ggplot(aes(x = wpre, y = wgain)) + 
  geom_point() + 
  theme_bw() + 
  xlab("Weight pre-pregnancy (lbs)") + 
  ylab("Weight gain (lbs)") + 
  ggtitle("Scatterplot of pre-pregnancy weight and weight gain")
```

\textcolor{blue}{These does not seem to be a strong association between pre-pregnancy weight and weight gain. There is no clear trend in the data points. Qualitatively, we have little evidence of a relationship.}

e. Reproduce your plot from part (e), but this time use `ggplot2` aesthetics to differentiate individuals with `parity = 0` (i.e. those who have had no previous births) and `parity > 0` (i.e., those who have had previous births). You will want to create a new variable for this problem, and then use color or shape to indicate which data points have previous births and which do not. (2 points)

```{r 1e}
births <- births %>% mutate(prev = ifelse(parity > 0, "yes", "no"))
births %>% ggplot(aes(x = wpre, y = wgain)) + 
  geom_point(aes(color = prev), alpha = 0.5) + # the alpha option makes the points slightly transparent
  theme_bw() + 
  xlab("Weight pre-pregnancy (lbs)") + 
  ylab("Weight gain (lbs)") + 
  labs(color = "Previous births") + 
  ggtitle("Scatterplot of pre-pregnancy weight and weight gain, by parity")
```

\newpage

## 2. Standard errors and confidence intervals

a. What is the mean birth parent age in the `births` dataset? Give an estimate of the standard error of the sample mean. Construct a 95\% confidence interval for the population mean age "by hand" (i.e., using `R` for calculations, but not using the `confint` function). Interpret this confidence interval. (2 points)

```{r}
mean_age <- births %>% pull(age) %>% mean()
se_est <- births %>% pull(age) %>% sd()/sqrt(nrow(births))
ci_l <- mean_age - 1.96*se_est
ci_u <- mean_age + 1.96*se_est
```

\textcolor{blue}{The mean birth parent age is `r round(mean_age, 1)` years. The standard error of the sample mean is $\sqrt{\sigma^2/n}$, where $\sigma^2$ is the variance of birth parent age in the population. We can estimate this using the sample variance, yielding a standard error estimate of `r round(se_est, 2)`. The form of a 95\% confidence interval is est $\pm 1.96\times$SE. If we plug in our sample mean and standard error estimate, we get a confidence interval of  (`r round(ci_l, 1)`, `r round(ci_u, 1)`).} The observed data would not be considered unusual if the true population mean age were between `r round(ci_l, 1)` years and `r round(ci_u, 1)` years. 

b. Suppose we took another sample of size 2500 from the live births in King County. If the standard deviation of birth parent age were larger in this sample, how would a 95\% confidence interval for mean age differ from what you determined in part (a)? Give one sentence explaining your rationale. (2 point)

\textcolor{blue}{The confidence interval from the new sample would be wider. The width of the confidence interval is determined by (1) the confidence level, which has not changed; (2) the sample size, which has not changed; and (3) the sample standard deviation, is larger in the new sample. A larger sample standard deviation leads to a larger standard error estimate, and a correspondigly wider confidence interval.}

c. What is the proportion of births assigned female sex? Estimate the standard error of the sample proportion and costruct a 95\% confidence interval "by hand" for the population proportion (i.e., using `R` for calculations, but not using the `confint` function). Interpret this confidence interval. (2 points)

```{r}
births <- births %>% mutate(sexn = ifelse(sex == "F", 1, 0))
p_female <- births %>% pull(sexn) %>% mean()
se_est <- sqrt(p_female*(1 - p_female)/nrow(births))
ci_l <- p_female - 1.96*se_est
ci_u <- p_female + 1.96*se_est
```

\textcolor{blue}{The proportion of births assigned female sex in this sample is `r round(p_female, 2)`. The standard error of a sample proportion is $\sqrt{p(1-p)/n}$, where $p$ is the true population proportion. We can estimate this using our sample proportion, yielding a standard error estimate of `r round(se_est, 2)`. The corresponding 95\% confidence interval is (`r round(ci_l, 2)`, `r round(ci_u, 2)`). The observed data would not be considered unusual if the true proportion assigned female sex were between `r round(ci_l, 2)` and `r round(ci_u, 2)`.}

\newpage

## 3. Study design

For each of the following scientific questions, propose an appropriate study design. (You may use the same design at most twice.) Provide a one-sentence explanation of your reasoning. 

a. Among people who are already heavy smokers, can running reduce the effect of smoking on incidence of lung cancer? (2 point)

\color{blue} **Possible Answers:**

- **Randomized controlled trial:** enroll heavy smokers without lung cancer and randomize either to a particular exercise program or to no intervention (participants can do whatever they would do anyway). Follow to see if/when subjects get lung cancer. 
    - *This would establish causality! If you can get funding for it, this would be our preference. Otherwise, a cohort, case-control, or cross-sectional study would be reasonable.*
- Case-control study: enroll people with lung cancer and people without lung cancer who have been heavy smokers at some point in the past [population of interest]. Compare how many runners are in each group. 
    - *This study design is attractive for rare outcomes like cancer (although among heavy smokers lung cancer is not particularly rare), but would show association rather than causation; also, it may be difficult to find the people in your population of interest (i.e., heavy smokers at some particular point in the past).* 
- Cohort study: enroll heavy smokers without lung cancer, follow for a set period of time and see whether they get lung cancer, and compare risk of lung cancer between runners and non-runners.
    - *Easier to define population of interest (current smokers), but would need a large sample to see many cases of the disease outcome over a reasonable period of time.*
- Cross-sectional study: enroll a random sample of heavy smokers. Ask them whether they run or not (either in the present or in the past) and whether they have lung cancer or not. 
    - *Like a cohort study, this study design is likely to be less efficient than case-control. You'd probably need a large sample, but this would be a valid way to study association.* 

\color{black}

b.	How does risk of severe influenza differ between those who received a flu vaccine and those who did not? (2 point)

\color{blue} **Possible Answers:**

- **Cohort study:** enroll vaccinated and unvaccinated individuals and see how many get the flu during a set period of time [e.g., one flu season]. 
    - *This would be our preference, since flu is a common and fairly rapidly-occurring outcome, so it's not terribly difficult or expensive to watch people and record whether they get the flu. Depending on how many of those cases are classified as severe, though, this could be somewhat inefficient.* 
- We probably can't use an RCT here for ethical reasons. It's well-established that flu vaccines work (although efficacy varies from year to year), so randomizing people to not receive the vaccine would be unethical. 
- Case-control study: enroll people with and without the flu, then see what proportion of each are vaccinated/unvaccinated.
- Cross-sectional study: enroll people randomly and see what proportion of vaccinated vs. unvaccinated have the flu. 

\color{black}

c. Is a common genetic mutation associated with colon cancer? (2 point)

\color{blue} **Possible Answers:**

- A **case-control study** is your best bet. This gives higher efficiency for the rare outcome, and you still know the temporal sequence of exposure $\rightarrow$ outcome because this genetic mutation is present at birth. 
- We can't use an RCT here, since you can't randomize someone to have or not have a genetic mutation. 
- It doesn't really make sense to do a cohort study here; we already know the temporal sequence of genetics $\rightarrow$ colon cancer, so there's no particular advantage and the study would need to be much larger and more expensive.  
- It doesn't really make sense to do a cross-sectional survey here, since it would be highly inefficient for a rare outcome like colon cancer. The context in which you'd do a cross-sectional study for genetic association analysis is when you are planning to take a very large sample and look at a wide variety of genetic markers and outcomes/diseases.  

\color{black}