---
title: "Homework 6: Key"
subtitle: "BIOST 311"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = "hide")
```


## Instructions

This assignment should be submitted on Canvas as a .pdf by 11:59pm on Friday, May 13th. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code (if applicable) in the same document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

Each homework assignment is worth 20 points total. Problems are graded for correctness, according to the number of points given in the problem statement. 

## 1. Summarizing binary variables

For this question, we will be using the `wcgs` dataset to explore ways to summarize binary variables. Recall that in this study, investigators were interested in knowing whether Type A behavior is associated with coronary heart disease. In this problem, we will compute various statistics that can help us answer this scientific question.


```{r preliminaries}
library(readr)
library(tidyverse)
library(ggplot2)
library(knitr)
```

a. Load the `wcgs` dataset (0 points)

```{r 1a}
wcgs <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/wcgs.csv"))
```

b. Provide relevant descriptive statistics for our outcome variable, `chd` (1 point).

```{r}
wcgs %>% pull(chd) %>% table()
wcgs %>% pull(chd) %>% table() %>% prop.table()
```

\textcolor{blue}{There are 257 (8.1\%) people in the study who have coronary heart disease, and 2897 (91.9\%) people who do not have coronary heart disease.}

c. Provide relevant descriptive statistics for our predictor of interest, `tabp` (1 point).

```{r}
wcgs %>% pull(tabp) %>% table()
wcgs %>% pull(tabp) %>% table %>% prop.table
```

\textcolor{blue}{There are 1589 (50.4\%) people in the study who have Type A behavior, and 1565 (49.6\%) people who do not have Type A behavior.}

d. Report the proportion of people who have coronary heart disease among those with Type A behavior, and the proportion of people who have coronary heart disease among those without Type A behavior. Your answer should be in complete sentences, not only reporting numbers (1 point). 

```{r}
wcgs %>% filter(tabp == 1) %>% pull(chd) %>% mean()
wcgs %>% filter(tabp == 0) %>% pull(chd) %>% mean()
```

\textcolor{blue}{11.2\% of people who have Type A behavior have coronary heart disease, and 5.0\% of people who do not have Type A behavior have coronary heart disease.}

e. Report and interpret the risk difference of coronary heart disease between those with Type A behavior and those without. If the risk of coronary heart disease between those with Type A behavior and those without was the same, what would you expect the risk difference to be? (1 point)

```{r}
risk_typea <- wcgs %>% filter(tabp == 1) %>% pull(chd) %>% mean()
risk_nottypea <- wcgs %>% filter(tabp == 0) %>% pull(chd) %>% mean()

risk_typea - risk_nottypea
```

\textcolor{blue}{We estimate that the difference in probability of having coronary heart disease between those with Type A behavior and those who do not have Type A behavior is 6.2\%. If the risk of coronary heart disease between those with Type A behavior and those without were the same, we would expect the risk difference to be 0.}

f. Report and interpret the relative risk of coronary heart disease between those with Type A behavior and those without. If the risk of coronary heart disease between those with Type A behavior and those without were the same, what would you expect the relative risk to be? (1 point)

```{r}
risk_typea / risk_nottypea
```

\textcolor{blue}{We estimate that the risk of coronary heart disease among those with Type A behavior is 2.22 times as large as the risk of coronary heart disease among those without Type A behavior. If the risk of coronary heart disease between those with Type A behavior and those without was the same, we would expect the relative risk to be 1.}

g. Report the odds of having coronary heart disease among those with Type A behavior, and the odds of having coronary heart disease among those without Type A behavior. Your answer should be in complete sentences, not only reporting numbers. (1 point)

```{r}
odds_typea <- risk_typea / (1 - risk_typea)
odds_nottypea <- risk_nottypea / (1 - risk_nottypea)
odds_typea
odds_nottypea
```

\textcolor{blue}{We estimate that the odds of coronary heart disease among those with Type A behavior are 0.126, and the odds of coronary heart disease among those without Type A behavior are 0.053.}

h. Report and interpret the ratio of odds of coronary heart disease, comparing people with and without a Type A behavior. If the odds of coronary heart disease between those with Type A behavior and those without were the same, what would you expect the odds ratio to be? (1 point)

```{r}
odds_typea / odds_nottypea
```

\textcolor{blue}{We estimate that the odds of coronary heart disease among those with Type A behavior are 2.4 as large as the odds of coronary heart disease among those without Type A behavior. If the odds of coronary heart disease between those with Type A behavior and those without were the same, we would expect the odds ratio to be 1.}

## 2. Linear regression with binary outcomes

a. Recall that in this study, investigators were interested in knowing whether Type A behavior is associated with coronary heart disease. Translate this scientific question into a statistical question that can be answered using simple *linear* regression. (1 point)

\textcolor{blue}{Is there a difference in probability of coronary heart disease between those with and without Type A behavior?}

b. Perform a linear regression analysis to answer the statistical question from part (a). Be sure to include a statistic with appropriate uncertainty quantification, as well as a hypothesis test. Be sure to state the null and alternative hypotheses. You should interpret your results; don’t just give numerical output. (2 points)

```{r}
mod <- lm(data = wcgs, chd ~ tabp)
summary(mod)
confint(mod)
```

\textcolor{blue}{The estimated difference in probability of coronary heart disease between those with and without Type A behavior is 6.2\% (95\% CI 4.3\% - 8.1\%), with those with Type A behavior having higher probability of CHD on average. This estimated difference in probability would not be unusual if the true difference were between 4.3\% and 8.1\%. These data provide strong evidence of a difference in probability of coronary heart disease between those with and without Type A behavior (p < 0.001) based on a test of the null hypothesis that the regression slope is 0, versus an alternative that the regression slope is not equal to 0.}

c. Compare your estimate from part (b) to the risk difference you estimated in Question 1 part (e). What do you notice? (1 point)

\textcolor{blue}{The estimate of the risk difference from the linear regression model and from calculating the risk difference by hand in Question 1 part (e) are identical.}

## 3. Simple logistic regression

a. Recall that in this study, they were interested in knowing whether Type A behavior is associated with coronary heart disease. Translate this scientific question into a statistical question that can be answered using simple *logistic* regression. (1 point)

\textcolor{blue}{Is the ratio of odds of coronary heart disease between those with and without Type A behavior different from 1?}

b. Perform a logistic regression analysis to answer the statistical question from part (a). Be sure to include a statistic with appropriate uncertainty quantification, as well as a hypothesis test. Be sure to state the null and alternative hypotheses. You should interpret your results; don’t just give numerical output. (2 points)

```{r}
mod <- glm(data = wcgs, chd ~ tabp, family = binomial)
summary(mod)
exp(coef(mod))
exp(confint(mod))
```

\textcolor{blue}{We estimate that the odds of coronary heart disease among those with Type A behavior are 2.37 times as large as the odds of coronary heart disease among those without Type A behavior (95\% CI 1.81 - 3.14). This estimated ratio of odds would not be unusual if the true odds ratio were between 1.81 and 3.14. These data provide strong evidence of a ratio of odds of coronary heart disease between those with and without Type A behavior different than 1 (p < 0.001) based on a test of the null hypothesis that the regression slope is 0, versus an alternative that the regression slope is not equal to 0.}

c. Compare your estimate from part (b) to the odds ratio you estimated in Question 1 part (h). What do you notice? (1 point)

\textcolor{blue}{The estimate of the odds ratio from the logistic regression model and from calculating the odds ratio by hand in Question 1 part (h) are identical.}

## 4. Linear and logistic regression with a binary outcome and a quantitative predictor

a. Suppose that we are interested in the association between coronary heart disease and blood pressure, measured in mmHg (millimeters of mercury). Studies have shown that a greater risk of stroke and heart disease is related to higher systolic pressure versus higher diastolic pressure. As such, we are particularly interested in the association between coronary heart disease and *systolic* blood pressure (`sbp` in our dataset). Translate this scientific question into *two* statistical questions: one that can be answered using simple *linear* regression, and one that can be answered using simple *logistic* regression. (1 point)

\textcolor{blue}{1. Is there a difference in probability of coronary heart disease, comparing two groups that differ by one mmHg in systolic blood pressure?}

\textcolor{blue}{2. Is the ratio of odds of coronary heart disease different from 1, comparing two groups that differ by one mmHg in systolic blood pressure?}

b. Perform a linear regression analysis to answer one of your statistical questions from part (a). Be sure to include a statistic with appropriate uncertainty quantification, as well as a hypothesis test. Be sure to state the null and alternative hypotheses. You should interpret your results; don’t just give numerical output. (2 points)

```{r}
mod <- lm(data = wcgs, chd ~ sbp)
summary(mod)
confint(mod)
```

\textcolor{blue}{The estimated difference in probability of coronary heart disease comparing two groups that differ by one mmHg in systolic blood pressure is 0.24\% (95\% CI 0.18\% - 0.30\%), with those with higher systolic blood pressure having higher probability of CHD on average. This estimated difference in probability would not be unusual if the true difference were between 0.18\% and 0.30\%. These data provide strong evidence of a difference in probability of coronary heart disease between groups differing by one mmHg in systolic blood pressure (p < 0.001) based on a test of the null hypothesis that the regression slope is 0, versus an alternative that the regression slope
is not equal to 0.}

c. Perform a logistic regression analysis to answer your other statistical question from part (a). Be sure to include a statistic with appropriate uncertainty quantification, as well as a hypothesis test. Be sure to state the null and alternative hypotheses. You should interpret your results; don’t just give numerical output. (2 points)

```{r}
mod <- glm(data = wcgs, chd ~ sbp, family = binomial)
summary(mod)
exp(coef(mod))
exp(confint(mod))
```

\textcolor{blue}{We estimate that the odds of coronary heart disease are 1.03 times as large for those with one mmHg higher systolic blood pressure (95\% CI 1.02 - 1.03). This estimated ratio of odds would not be unusual if the true odds ratio were between 1.02 and 1.03. These data provide strong evidence of a ratio of odds of coronary heart disease comparing two groups that differ by one mmHg in systolic blood pressure that is different than 1 (p < 0.001) based on a test of the null hypothesis that the regression slope is 0, versus an alternative that the regression slope is not equal to 0.}
