---
title: "Homework"
author: "Taylor Okonek and Charlie Wolock"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

## Instructions

This assignment should be submitted on Canvas as a .pdf by 12:30pm on the due date. You are welcome to work together on the homework; however, final solutions should be written up individually. 

Make sure you update the document header to reflect the assignment number (e.g., "Homework 1"), your name ("Taylor Okonek and Charlie Wolock" > "Your Name"). Unedited software output in the body of your assignment is not acceptable. Please round numerical values to 2--3 significant digits, and include your R code (if applicable) in the same document. See the "How to Write a Proper Homework" document on Canvas for additional guidelines.  

Each homework assignment is worth 20 points total. Problems are graded for correctness, according to the number of points given in the problem statement. 

## 1. (XX points)

Type your answer to this problem here.

```{r code-for-prob-1, echo = TRUE, results = "hide"}
## This is a sample code chunk! 
## Fill this in with code executing the R commands you need for this problem
## Comment your code so that you (and us, if necessary) know what you're doing
## The options on this code chunk will include the code in your finished file, but not the raw output
```

## 2. (YY points)

Type your answer to this problem here.

```{r code-for-prob-2, echo = TRUE, results = "hide"}
## Fill this in with code executing the R commands you need for this problem
## Comment your code so that you (and us, if necessary) know what you're doing
```
