---
title: "Logistic Regression"
subtitle: "BIOST 311, Discussion Section Week 7: Key"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = "hide")
```

# Introduction

In this discussion section, we'll walk through a logistic regression analysis. We will use the `wcgs` dataset that we have been using in class for the Chapter 2.

The dataset is called `wcgs.csv`, and is available on the Canvas site on the Week 6 page, or on the course github site. The code is provided below to load the dataset directly from github, but you may download it to your personal machine and load it in that way if you prefer.

```{r}
library(tidyverse)
library(readr)
wcgs <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/wcgs.csv"))
```

Suppose we are interested in the association between high blood pressure and cholesterol. When arteries build up with plaque from cholesterol, the heart needs to work harder to pump blood through them. Therefore, we have reason to believe that higher levels of cholesterol are associated with high blood pressure.

For the purposes of this analysis, we will be defining someone with high blood pressure (a binary indicator) as someone who has a diastolic blood pressure greater than 90, and a systolic blood pressure greater than 140. This is a relatively common definition of high blood pressure, or hypertension. 

# Exploratory plot and set-up

1. Create a binary indicator for high blood pressure in the `wcgs` dataset. Call this new variable `highbp`. Provide descriptive summary statistics for the variable `highbp`.

```{r}
wcgs <- wcgs %>%
  mutate(highbp = ifelse(dbp > 90 & sbp > 140, 1, 0))

wcgs %>% pull(highbp) %>% table()
wcgs %>% pull(highbp) %>% table() %>% prop.table()
```

\textcolor{blue}{313 (9.9\%) individuals in our dataset had high blood pressure, and 2841 (90.1\%) individuals did not have high blood pressure.}

2. Create an exploratory plot that describes the relationship between high blood pressure and cholesterol. How would you describe the relationship between these two variables?

```{r}
wcgs %>%
  mutate(`High BP` = factor(highbp)) %>% 
  ggplot(aes(`High BP`, chol, fill = `High BP`)) +
  geom_boxplot() +
  xlab("High Blood Pressure") +
  ylab("Cholesterol (mg/dL)") +
  ggtitle("High Blood Pressure vs. Cholesterol")
```

\textcolor{blue}{The medians and IQRs of cholesterol for those with and without high blood pressure are relatively similar, though the spread of cholesterol seems to be slightly higher those without high blood pressure.}

# Simple logistic regression

3. Fit a simple logistic regression model with high blood pressure as the outcome and cholesterol as the predictor of interest. Interpret the results of your analysis using an odds ratio in context, and conduct a hypothesis test. Make sure to report uncertainty in the form of a confidence interval in addition to a point estimate.

```{r}
mod <- glm(data = wcgs, highbp ~ chol, family = binomial)
summary(mod)
exp(coef(mod))
exp(confint(mod))
```

\textcolor{blue}{We estimate that the odds of high blood pressure are 1.004 times as large for individuals with one mg/dL higher cholesterol. This observed odds ratio would not be surprising if the true odds ratio were between 1.002 and 1.007. We reject the null hypothesis that the ratio of odds is 1 (p < 0.05), where the alternative hypothesis is that the ratio of odds is not equal to 1. We conclude that we have statistically significant scientific evidence of an association between high blood pressure and cholesterol.}

# Multiple logistic regression

We have reason to believe that weight is a confounding variable in the relationship between cholesterol and high blood pressure. Having higher weight puts you at higher risk for high blood pressure (a causal relationship), and similarly can lead to increased cholesterol levels. 

4. Fit a multiple logistic regression model with high blood pressure as the outcome, cholesterol as the predictor of interest, and weight as an additional variable in your model. Interpret the results of your analysis using an odds ratio in context, and conduct a hypothesis test. Make sure to report uncertainty in the form of a confidence interval in addition to a point estimate.

```{r}
mod <- glm(data = wcgs, highbp ~ chol + weight, family = binomial)
summary(mod)
exp(coef(mod))
exp(confint(mod))
```

\textcolor{blue}{We estimate that the odds of high blood pressure are 1.004 times as large for individuals with one mg/dL higher cholesterol who have the same weight. This observed odds ratio would not be surprising if the true odds ratio were between 1.002 and 1.007. We reject the null hypothesis that the ratio of odds is 1 (p < 0.05), where the alternative hypothesis is that the ratio of odds is not equal to 1. We conclude that we have statistically significant scientific evidence of an association between high blood pressure and cholesterol for individuals of the same weight.}



