---
title: "Survival Analysis"
subtitle: "BIOST 311, Discussion Section Week 10: Key"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = "hide")
```

# Introduction

In this discussion section, we'll walk through a survival analysis problem. We will use *mayo* dataset that we have used in class as well as Homework 8. As a reminder, the mayo dataset comes from a randomized controlled trial of d-penicillamine for treatment of primary biliary cirrhosis (PBC). PBC is a liver disease that affects the bile ducts, and can eventually lead to severe liver damage, called cirrhosis. There’s some evidence that PBC has an autoimmune component, and d-penicillamine was considered due to its possible effect on autoimmune processes.

In this trial, 311 patients were randomized to receive either treatment or placebo. Baseline health information was collected (mostly about baseline liver function), and patients were followed until death or censoring (due to either administrative censoring or dropout).

```{r}
library(tidyverse)
library(readr)
library(survival)
library(survminer)
mayo <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/mayo.csv"))
```

# Exploratory plots

1. Make two descriptive plots for the `cholesterol` and `age` variables in the *mayo* dataset, respectively. Provide descriptive summaries for these two variables as well, including the range of values observed for each variable.

```{r}
mayo %>%
  ggplot(aes(cholest)) +
  geom_histogram(bins = 7, col = "black") +
  labs(x = "Serum cholesterol level (mg/dL)", y = "Count", title = "Distribution of Serum cholesterol level")

summary(mayo %>% pull(cholest))

mayo %>%
  ggplot(aes(age)) +
  geom_histogram(bins = 15, col = "black") +
  labs(x = "Age (years)", y = "Count", title = "Distribution of Age at randomization")

summary(mayo %>% pull(age))
```

\textcolor{blue}{The distribution of serum cholesterol levels is very right-skewed, with the minumum being -9 mg/dL, and the maximum being 1775 mg/dL. The individuals with -9 mg/dL recorded are likely missing values. The mean serum cholesterol level in our dataset is 336 mg/dL, which is very high.}

\textcolor{blue}{The distribution of age is roughly symmetric, with the minumum age being 26 years, and the maximum being 78 years. The mean age at randomization in our dataset is 50 mg/dL.}

2. You should have noticed a few individuals with cholesterol levels that do not make sense. Sometimes, rather than recording missing values as "NA" in datasets, people will record missing values with "-9", "-99", or "-999". This is typically reported in the metadata for a given dataset, but not for this one! Create a new dataset that includes only observations with cholesterol levels that make sense (aka cholesterol levels that are not missing). *This is the dataset we will use for all analyses and plots from here on.*

```{r}
mayo_new <- mayo %>% filter(cholest != -9)
```

3. Suppose we are interested in the association between survival time and cholesterol. Create a new, binary variable called `cholest_high` that is either 0 if `cholest` is less than 300, or is 1 if `cholest` is greater than or equal to 300. Plot Kaplan-Meier survival curves for all individuals in the dataset, stratified by `cholest_high`. Include a legend in your plot and confidence bands for each strata. Based on your plot, what do you notice about the relationship between time to death and cholesterol? 

```{r}
mayo_new <- mayo_new %>%
  mutate(cholest_high = cholest >= 300)

fit <- survfit(Surv(time = obstime, event = status) ~ cholest_high,
               data = mayo_new)
ggsurvplot(fit, data = mayo_new,
           censor.shape = "|",
           legend.labs = c("Cholesterol < 300", "Cholesterol >= 300"),
           censor.size = 4,
           conf.int = TRUE,
           legend.title = element_blank()) +
  labs(x = "Time (days)",
       title = "Kaplan-Meier survival curves by High Cholesterol") 
```

\textcolor{blue}{Prior to 1000 days, those with high cholesterol seem to have a higher probability of survival than those with low cholesterol. After 1000 days, those with low cholesterol seem to have a higher probability of survival than those with high cholesterol.}

4. If we were to fit a Cox proportional hazards regression to estimate the association between high cholesterol and time to death, would you have any concerns about the assumptions of the Cox proportional hazards model holding? You may wish to use the plot you made in Question 3 to justify your answer.

\textcolor{blue}{The proportional hazards assumption would certainly not hold, since the survival curves for high and low cholesterol groups cross.}

5. Suppose we are also interested in the association between survival time and age. Create a new, binary variable called `age_high` that is either 0 if `age` is less than 50, or is 1 if `age` is greater than or equal to 50. Plot Kaplan-Meier survival curves for all individuals in the dataset, stratified by `age_high`. Include a legend in your plot and confidence bands for each strata. Based on your plot, what do you notice about the relationship between time to death and age? 

```{r}
mayo_new <- mayo_new %>%
  mutate(age_high = age >= 50)

fit <- survfit(Surv(time = obstime, event = status) ~ age_high,
               data = mayo_new)
ggsurvplot(fit, data = mayo_new,
           censor.shape = "|",
           legend.labs = c("Age < 50", "Age >= 50"),
           censor.size = 4,
           conf.int = TRUE,
           legend.title = element_blank()) +
  labs(x = "Time (days)",
       title = "Kaplan-Meier survival curves by Age") 
```

\textcolor{blue}{Perhaps unsurprisingly, those who are younger seem to have a higher probability of survival than those who are older.}

# Cox proportional hazards regression

5. Perform a Cox proportional hazards regression analysis to assess whether cholesterol is associated with survival time. Interpret the output of your model; don't just give numerical output. Be sure to include a statistic quantifying the association with appropriate uncertainty quantification, as well as hypothesis tests. Be sure to state the null and alternative hypotheses. 

```{r}
mod <- coxph(Surv(time = obstime, event = status) ~ cholest, data = mayo_new)
summary(mod)
```

\textcolor{blue}{We estimate the hazard ratio of death comparing groups one mg/dL apart in cholesterol is 1.001, with the group with higher cholesterol having the higher hazard. The data would not be surprising if the true hazard ratio were between 1.001 and 1.002. We reject the null hypothesis of a hazard ratio of 1 against the alternative that the hazard ratio is not equal to 1 (p < 0.001). There is significant evidence of an association between cholesterol and instantaneous risk of death.}

6. Perform a Cox proportional hazards regression analysis to assess whether age modifies the association between cholesterol and survival time. Interpret the output of your model for groups who are 40 years old, and groups who are 60 years old; don't just give numerical output. Be sure to include a statistic quantifying the association with appropriate uncertainty quantification, as well as hypothesis tests. Be sure to state the null and alternative hypotheses. 

```{r}
mod <- coxph(Surv(time = obstime, event = status) ~ cholest + age + cholest:age, data = mayo_new)
summary(mod)

exp(coef(mod)[1] + coef(mod)[3] * 40)
exp(coef(mod)[1] + coef(mod)[3] * 60)
```


\textcolor{blue}{We estimate the hazard ratio of death comparing groups one mg/dL apart in cholesterol is 1.0017 for those who are 40 years old, with the group with higher cholesterol having the higher hazard. We estimate the hazard ratio of death comparing groups one mg/dL apart in cholesterol is 1.0009 for those who are 60 years old, with the group with higher cholesterol having the higher hazard. The data would not be surprising if the true ratio of hazards ratios were between 0.9999 and 1.000. We fail to reject the null hypothesis of a ratio of hazard ratios of 1 against the alternative that the ratio of hazard ratios is not equal to 1 (p = 0.291). We do not have statistically significant evidence of age modifiying the association between cholesterol and instantaneous risk of death.}
