---
title: "Survival Analysis"
subtitle: "BIOST 311, Discussion Section Week 10"
author: "Taylor Okonek and Charlie Wolock"
date: \today
output: 
  pdf_document:
    toc: false
header-includes:
  - \usepackage{color}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(formatR)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(results = "hide")
```

# Introduction

In this discussion section, we'll walk through a survival analysis problem. We will use *mayo* dataset that we have used in class as well as Homework 8. As a reminder, the mayo dataset comes from a randomized controlled trial of d-penicillamine for treatment of primary biliary cirrhosis (PBC). PBC is a liver disease that affects the bile ducts, and can eventually lead to severe liver damage, called cirrhosis. There’s some evidence that PBC has an autoimmune component, and d-penicillamine was considered due to its possible effect on autoimmune processes.

In this trial, 311 patients were randomized to receive either treatment or placebo. Baseline health information was collected (mostly about baseline liver function), and patients were followed until death or censoring (due to either administrative censoring or dropout).

```{r}
library(tidyverse)
library(readr)
library(survival)
library(survminer)
mayo <- read_csv(url("https://raw.githubusercontent.com/cwolock/BIOST311/main/Datasets/mayo.csv"))
```

# Exploratory plots

1. Make two descriptive plots for the `cholesterol` and `age` variables in the *mayo* dataset, respectively. Provide descriptive summaries for these two variables as well, including the range of values observed for each variable.

2. You should have noticed a few individuals with cholesterol levels that do not make sense. Sometimes, rather than recording missing values as "NA" in datasets, people will record missing values with "-9", "-99", or "-999". This is typically reported in the metadata for a given dataset, but not for this one! Create a new dataset that includes only observations with cholesterol levels that make sense (aka cholesterol levels that are not missing). *This is the dataset we will use for all analyses and plots from here on.*

3. Suppose we are interested in the association between survival time and cholesterol. Create a new, binary variable called `cholest_high` that is either 0 if `cholest` is less than 300, or is 1 if `cholest` is greater than or equal to 300. Plot Kaplan-Meier survival curves for all individuals in the dataset, stratified by `cholest_high`. Include a legend in your plot and confidence bands for each strata. Based on your plot, what do you notice about the relationship between time to death and cholesterol? 

4. If we were to fit a Cox proportional hazards regression to estimate the association between high cholesterol and time to death, would you have any concerns about the assumptions of the Cox proportional hazards model holding? You may wish to use the plot you made in Question 3 to justify your answer.

5. Suppose we are also interested in the association between survival time and age. Create a new, binary variable called `age_high` that is either 0 if `age` is less than 50, or is 1 if `age` is greater than or equal to 50. Plot Kaplan-Meier survival curves for all individuals in the dataset, stratified by `age_high`. Include a legend in your plot and confidence bands for each strata. Based on your plot, what do you notice about the relationship between time to death and age? 

# Cox proportional hazards regression

5. Perform a Cox proportional hazards regression analysis to assess whether cholesterol is associated with survival time. Interpret the output of your model; don't just give numerical output. Be sure to include a statistic quantifying the association with appropriate uncertainty quantification, as well as hypothesis tests. Be sure to state the null and alternative hypotheses. 

6. Perform a Cox proportional hazards regression analysis to assess whether age modifies the association between cholesterol and survival time. Interpret the output of your model for groups who are 40 years old, and groups who are 60 years old; don't just give numerical output. Be sure to include a statistic quantifying the association with appropriate uncertainty quantification, as well as hypothesis tests. Be sure to state the null and alternative hypotheses. 


